---
title: "Extracting Data into R"
author: "Brandon Monier"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
    BiocStyle::html_document:
      toc: true
      toc_float: true
      fig_caption: true
      number_sections: false
vignette: >
    %\VignetteIndexEntry{Extracting Data into R}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
    %\usepackage{float}
---

```{r setup, include=FALSE}
library(rTASSEL)

knitr::opts_chunk$set(
    fig.path='figure/graphics-',
    cache.path='cache/graphics-',
    fig.align='center',
    external=TRUE,
    echo=TRUE,
    warning=FALSE
    # fig.pos="H"
)
```


# Introduction

## Where the data actually lives
An `rTASSEL` object is a handle, not a copy. When you call
`readGenotype()` or `readPhenotype()`, the table itself is built inside the
Java Virtual Machine and what comes back to R is a small S4 object holding
a pointer to it:

```{r, eval=TRUE, echo=TRUE}
genoPath <- system.file("extdata", "mdp_genotype.hmp.txt", package = "rTASSEL")

myGT <- readGenotype(genoPath)

slotNames(myGT)
```

The `jRefObj` slot is the pointer, and the other two record what it points
at. Nothing crosses back into R until you ask it to, which is what keeps a
genotype table of any size cheap to filter and pass around.

This vignette is about asking. Every way of turning a TASSEL object into an
ordinary R object - a `character` vector, a `tibble`, a `matrix` - is
collected here, grouped by the class you start from.

**Prior knowledge about `rTASSEL` data structures is needed, so please refer
to the vignette *Getting Started with rTASSEL* if you are new to this
package!**

*Note:* We will be using the base R pipe operator `|>` throughout. More
information about this can be found [here](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html).


## The whole surface at a glance
Two kinds of function appear below. **Accessors** read one piece of an
object and hand back an R object, and are named for what they return
(`taxaList()`, `positionList()`, `attributeData()`). **Coercion** methods
are the base R generics `as.matrix()` and `as.data.frame()`, which return
the object's principal table.

| Class                  | Function              | Returns                                    |
|------------------------|-----------------------|--------------------------------------------|
| `TasselGenotype`       | `taxaList()`          | `character` of taxa IDs                    |
|                        | `positionList()`      | `tibble` of marker positions               |
|                        | `seqnames()`          | `character` of chromosome IDs               |
|                        | `granges()`           | `GRanges`, one width-1 range per marker    |
|                        | `siteSummary()`       | `tibble`, one row per site                 |
|                        | `taxaSummary()`       | `tibble`, one row per taxon                |
|                        | `as.matrix()`         | `integer` dosage matrix, taxa by sites     |
|                        | `as.matrix(type = "allele")` | `character` matrix of calls         |
|                        | `as(x, "SummarizedExperiment")` | `SummarizedExperiment`, sites by taxa |
| `TasselNumericGenotype` | `as.matrix()`        | `numeric` matrix of reference probabilities |
| `TasselPhenotype`      | `as.data.frame()`     | `tibble`, one row per observation          |
|                        | `attributeData()`     | `tibble`, one row per column               |
|                        | `traitNames()`        | `character` of trait names                 |
|                        | `taxaList()`          | `character` of distinct taxa IDs           |
| `TasselGenomicDataset` | `genotype()`          | the joined `TasselGenotype`                |
|                        | `phenotype()`         | the joined `TasselPhenotype`               |
|                        | `as.matrix()`         | the genotype half, as above                |
|                        | `as.data.frame()`     | the phenotype half, as above               |
| `TasselDistanceMatrix` | `as.matrix()`         | `numeric` pairwise matrix                  |
|                        | `as.dist()`           | `dist`, the lower triangle                 |
|                        | `taxaList()`          | `character` of taxa IDs                    |
|                        | `dim()`, `nrow()`, `ncol()` | `numeric` dimensions                 |
|                        | `rownames()`, `colnames()`  | `character` of taxa IDs              |
| `AssociationResults`   | `tableReport()`       | `tibble`, or a named `list` of them        |
|                        | `reportNames()`       | `character` of report names                |
|                        | `traitNames()`        | `character` of modelled traits             |
|                        | `associationType()`   | `character` naming the model               |
| `PCAResults`           | `tableReport()`       | `tibble`, or a named `list` of them        |
|                        | `reportNames()`       | `character` of report names                |
| `LDResults`            | `tableReport()`       | `tibble` of pairwise LD statistics         |
|                        | `reportNames()`       | `character` of report names                |
| all of the above       | `javaRefObj()`        | the underlying `jobjRef`                   |

Every accessor that a `TasselGenotype` or a `TasselPhenotype` answers is
also answered by a `TasselGenomicDataset`, which passes it to whichever
half of itself holds the data.


## Types and missing values
Every table in this vignette comes across the same way, so two
conventions hold throughout. Column names lose their spaces: TASSEL's
`Minor Allele Frequency` is read as `Minor_Allele_Frequency`. And each
column type carries its own spelling of a missing value:

| Column type | Missing value |
|:------------|:--------------|
| `double`    | `NaN`         |
| `integer`   | `NA`          |
| `character` | `""`          |

`NaN` is what TASSEL itself records for a missing number, and `is.na()`
recognises it, so both numeric spellings answer the same test. The one
exception is the dosage matrix, which is built from a different Java
call and uses `NA` throughout.


## The example data
The examples below use the toy data sets shipped with `rTASSEL`: a hapmap
genotype table, and the phenotype file that carries all three non-taxa
TASSEL attribute types.

```{r, eval=TRUE, echo=TRUE}
phenoPath <- system.file("extdata", "mdp_phenotype.txt", package = "rTASSEL")

myPheno <- readPhenotype(phenoPath)

myDataset <- readGenomicDataset(myGT, myPheno)
```



# Genotype tables

## Taxa and positions
The two axes of a genotype table are read with `taxaList()` and
`positionList()`. The first returns a plain `character` vector of taxa IDs,
in the order TASSEL holds them:

```{r, eval=TRUE, echo=TRUE}
myGT |> taxaList() |> head()

myGT |> taxaList() |> length()
```

`positionList()` returns a `tibble` describing each marker:

```{r, eval=TRUE, echo=TRUE}
myGT |> positionList() |> head()
```

The `Site` column is TASSEL's own site index, which starts at 0. Indices
you hand to `rTASSEL` - `sites()`, `sitesWhere(siteIndex)`, and a bare
numeric second bracket index - are 1-based like every other index in R, so
`Site` is one less than the position a filter would count.

The chromosomes present, without the rest of the table, are given by
`seqnames()`:

```{r, eval=TRUE, echo=TRUE}
myGT |> seqnames()
```


## Site and taxa summaries
TASSEL's genotype summary is available for either axis. `siteSummary()`
computes one row per marker, with allele frequencies, missingness, and
heterozygosity among its columns:

```{r, eval=TRUE, echo=TRUE}
mySiteSummary <- myGT |> siteSummary()

dim(mySiteSummary)

mySiteSummary[, 1:8] |> head()
```

`taxaSummary()` is the same idea on the other axis, and is short enough to
show whole:

```{r, eval=TRUE, echo=TRUE}
myTaxaSummary <- myGT |> taxaSummary()

myTaxaSummary |> head()
```

Both are ordinary `tibble` objects, so they can go straight into a plot or
a model. Here is the distribution of minor allele frequencies:

```{r, eval=TRUE, echo=TRUE}
summary(mySiteSummary$Minor_Allele_Frequency)
```

These are the same quantities the filtering metadata is built from, under
TASSEL's longer names. A filter on `maf` reads
`Minor_Allele_Frequency` here, and `notMissing` is
`1 - Proportion_Missing`. See the *Filtering Genotype Tables* vignette for
the filtering spellings.


## Genotype calls as a matrix
`as.matrix()` brings the calls themselves into R as a **dosage matrix**,
where each cell is the number of alternate alleles a taxon carries at a
site. Taxa are rows and sites are columns:

```{r, eval=TRUE, echo=TRUE}
gtMat <- myGT |> as.matrix()

dim(gtMat)

gtMat[1:5, 1:4]
```

The result is an `integer` matrix with `NA` in place of missing calls, and
its dimnames come from `taxaList()` and the `Name` column of
`positionList()`:

```{r, eval=TRUE, echo=TRUE}
typeof(gtMat)

sum(is.na(gtMat))
```

This is the one extraction that is as large as the data, so it is worth
filtering first. A genotype table of any real size should be cut down to
the taxa and sites you need before it is materialized:

```{r, eval=TRUE, echo=TRUE}
myGT[, sitesWhere(chrom == "1" & maf >= 0.05)] |>
    as.matrix() |>
    dim()
```

Passing `type = "allele"` reads the same calls as TASSEL spells them
instead, which for nucleotide data means one IUPAC code per cell and
`"N"` for a missing call:

```{r, eval=TRUE, echo=TRUE}
alleleMat <- myGT[, sites(1:6)] |> as.matrix(type = "allele")

alleleMat[1:5, ]
```

The two readings have the same shape and dimnames, so either can be used
wherever a matrix of calls is wanted. The dosage matrix is the smaller of
the two and the one models expect; the allele matrix is the one to print.

Numeric genotype data is the exception. A `TasselNumericGenotype` holds
probabilities rather than discrete calls, so there is no dosage to report
and `as.matrix()` returns the reference probabilities instead:

```{r, eval=TRUE, echo=TRUE}
numGtPath <- system.file("extdata", "numeric_genotype.txt", package = "rTASSEL")

readGenotype(numGtPath) |> as.matrix()
```

TASSEL reads these one cell at a time, so unlike the dosage matrix this
one is worth filtering down to size first.


## Positions as a GRanges
`granges()` returns the marker positions as a
`GenomicRanges::GRanges` object, which is the form the rest of Bioconductor
expects. Each marker is a width-1 range, named for the marker:

```{r, eval=TRUE, echo=TRUE}
myRanges <- myGT |> granges()

myRanges |> head()
```

The rest of the `positionList()` columns come along as metadata columns
when asked for:

```{r, eval=TRUE, echo=TRUE}
myGT |> granges(use.mcols = TRUE) |> head()
```

A `GRanges` can be handed straight back to `region()` or `overlaps()` to
filter on, which is covered in *Filtering Genotype Tables*. This makes
`granges()` the way to cut one genotype table down to the markers of
another:

```{r, eval=TRUE, echo=TRUE}
myTargets <- myGT[, region("2", 20e6, 40e6)] |> granges()

myGT[, sitesWhere(overlaps(myTargets))]
```


## As a SummarizedExperiment
Where a Bioconductor container is wanted rather than a bare matrix,
`as()` assembles one from the dosage matrix, the positions, and the taxa:

```{r, eval=TRUE, echo=TRUE}
myGT[, sites(1:100)] |> as("SummarizedExperiment")
```

Note that a `SummarizedExperiment` puts features in rows and samples in
columns, which is the transpose of `as.matrix()`. The same works on a
`TasselGenomicDataset`, where it takes the genotype half.


## Writing to disk instead
When the destination is a file rather than an R object, there is no need to
go through a matrix. `exportGenotypeTable()` writes a genotype table
directly from Java in any of four formats:

```{r, eval=FALSE, echo=TRUE}
myGT |>
    exportGenotypeTable(
        file   = "my_genotypes.vcf",
        format = "vcf"
    )
```

The supported formats are `"vcf"`, `"hapmap"`, `"plink"`, and
`"flapjack"`.



# Phenotype data

## The trait table
A `TasselPhenotype` becomes a `tibble` with `as.data.frame()`, one row per
observation:

```{r, eval=TRUE, echo=TRUE}
myPhenoDF <- myPheno |> as.data.frame()

myPhenoDF |> head()
```

Column types are translated from each column's TASSEL attribute type:

| TASSEL attribute type | R type      |
|:----------------------|:------------|
| `taxa`                | `character` |
| `data`                | `numeric`   |
| `covariate`           | `numeric`   |
| `factor`              | `character` |

```{r, eval=TRUE, echo=TRUE}
sapply(myPhenoDF, class)
```

TASSEL records a missing value as `NaN`, which `is.na()` recognises:

```{r, eval=TRUE, echo=TRUE}
colSums(is.na(myPhenoDF))
```


## Attribute metadata
`as.data.frame()` returns the values alone. The attribute type of each
column is reported separately, by `attributeData()`:

```{r, eval=TRUE, echo=TRUE}
myPheno |> attributeData()
```

The `trait_type` column holds the TASSEL attribute type, `r_type` the R
type it is translated to, and `trait_attribute` the Java class backing it.
Because the two tables line up column for column, they can be used together
to pick out one kind of trait:

```{r, eval=TRUE, echo=TRUE}
myAttr <- myPheno |> attributeData()

covariates <- myAttr$trait_id[myAttr$trait_type == "covariate"]

myPhenoDF[covariates] |> head()
```


## Back the other way
The two tables are also what `readPhenotype()` needs to build a phenotype
from a data frame, so together they make the return trip:

```{r, eval=TRUE, echo=TRUE}
readPhenotype(myPhenoDF, attrTypes = myAttr)
```

The `attrTypes` argument takes either this spelling or the `col_id` and
`tassel_attr` columns documented in *Getting Started with rTASSEL*, so
metadata read off one phenotype can be used to build another.


## Names and taxa
The trait names, which is everything except the taxa column, come from
`traitNames()`:

```{r, eval=TRUE, echo=TRUE}
myPheno |> traitNames()
```

`taxaList()` means the same thing it does on a genotype table: the distinct
taxa. A phenotype may hold several observations of one taxon, so this is
generally shorter than the data frame is tall:

```{r, eval=TRUE, echo=TRUE}
myPheno |> taxaList() |> length()

myPheno |> as.data.frame() |> nrow()
```



# Genomic datasets

## The two components
A `TasselGenomicDataset` holds a genotype table and a phenotype joined on
their shared taxa. Either half can be pulled out whole, as the object it
would have been on its own:

```{r, eval=TRUE, echo=TRUE}
myDataset |> genotype()
```

```{r, eval=TRUE, echo=TRUE}
myDataset |> phenotype()
```

These are the *joined* tables rather than the ones handed to
`readGenomicDataset()`, so they carry only the taxa the join kept.

Every accessor from the two sections above also works on the dataset
directly, which passes it to whichever half can answer it:

```{r, eval=TRUE, echo=TRUE}
myDataset |> taxaList() |> length()

myDataset |> traitNames()
```


## Coercion
The two coercion methods each address the half they belong to.
`as.matrix()` returns the genotype calls and `as.data.frame()` the
phenotype table:

```{r, eval=TRUE, echo=TRUE}
myDataset |> as.matrix() |> dim()

myDataset |> as.data.frame() |> dim()
```


## Lining the two up
The matrix is indexed by taxon and the data frame by observation, so
joining them in R means joining on the taxa column. Since a taxon may be
observed more than once, the matrix rows are matched to the data frame
rows rather than the other way around:

```{r, eval=TRUE, echo=TRUE}
myMat <- myDataset |> as.matrix()
myDF  <- myDataset |> as.data.frame()

# One row of genotype calls per observation
myCalls <- myMat[myDF$Taxa, 1:4]

head(myCalls)
```

Both halves come out of the same joined object, so every taxon in the data
frame is guaranteed to be a row of the matrix.



# Distance and kinship matrices

## Matrix-like accessors
`kinshipMatrix()`, `distanceMatrix()`, and `readTasselDistanceMatrix()` all
return a `TasselDistanceMatrix`, which is another handle to a Java object.
It answers the base R generics you would ask of a `matrix`:

```{r, eval=TRUE, echo=TRUE}
myKin <- kinshipMatrix(myGT)

dim(myKin)

nrow(myKin)

myKin |> rownames() |> head()
```


## Coercion
`as.matrix()` brings the pairwise values across as an ordinary `numeric`
matrix:

```{r, eval=TRUE, echo=TRUE}
myKinMat <- myKin |> as.matrix()

class(myKinMat)

myKinMat[1:4, 1:4]
```

From there it is a base R matrix like any other, so it can go to
`heatmap()` or any modelling function that takes a relationship matrix.

Clustering functions want a `dist` rather than a matrix, which
`as.dist()` returns:

```{r, eval=TRUE, echo=TRUE}
myDistances <- distanceMatrix(myGT[, sites(1:500)])

myDistances |> as.dist() |> hclust()
```

A `dist` holds only the lower triangle, so the diagonal is dropped. Use
`as.matrix()` when the whole square is wanted.


## Back the other way
A pairwise R matrix can be handed to TASSEL with
`asTasselDistanceMatrix()`, which is the return trip:

```{r, eval=TRUE, echo=TRUE}
myKinMat |> asTasselDistanceMatrix()
```

The matrix must be square, and its row and column names must match.



# Analysis results

## Table reports
The analysis functions return objects that are already R-side: a list of
`tibble` objects, one per TASSEL table report, plus whatever metadata the
analysis produced. `tableReport()` reads them and `reportNames()` lists
what is there.

`tableReport()` with no arguments returns the default report for the
analysis that produced the object, which is the one you almost always want:

```{r, eval=TRUE, echo=TRUE}
myGLM <- assocModelFitter(
    tasObj     = myDataset,
    formula    = EarHT ~ .,
    fitMarkers = TRUE
)

myGLM |> reportNames()
```

```{r, eval=TRUE, echo=TRUE}
myGLM |> tableReport() |> head()
```

Naming a report returns that one instead:

```{r, eval=TRUE, echo=TRUE}
myGLM |> tableReport("GLM_Genotypes") |> head()
```

Passing `"ALL"` returns every report at once, as a named list:

```{r, eval=TRUE, echo=TRUE}
myGLM |> tableReport("ALL") |> names()
```


## Association results
Alongside the reports, an `AssociationResults` object knows which traits
were modelled and which model was fit:

```{r, eval=TRUE, echo=TRUE}
myGLM |> traitNames()

myGLM |> associationType()
```

`associationType()` is what the plotting functions read to decide how to
draw a result, and is worth checking when a pipeline handles several kinds
of model. The same three accessors work on every association subclass -
BLUE, GLM, MLM, Fast Association, and stepwise - with only the default
report differing:

```{r, eval=TRUE, echo=TRUE}
myBLUE <- assocModelFitter(
    tasObj     = myPheno,
    formula    = . ~ .,
    fitMarkers = FALSE
)

myBLUE |> reportNames()

myBLUE |> tableReport() |> head()
```


## PCA results
`pca()` returns a `PCAResults` object, which reads the same way. Its
default report is the principal components themselves:

```{r, eval=TRUE, echo=TRUE}
myPCA <- pca(myGT)

myPCA |> reportNames()

myPCA |> tableReport() |> head()
```

The eigenvalues and eigenvectors are the other two reports:

```{r, eval=TRUE, echo=TRUE}
myPCA |> tableReport("Eigenvalues_Datum") |> head()
```


## MDS results
`mds()` returns an `MDSResults` object, which follows the same pattern.
Its default report is the axes, one row per taxon:

```{r, eval=TRUE, echo=TRUE}
myDist <- distanceMatrix(myGT)

myMDS <- mds(myDist)

myMDS |> reportNames()

myMDS |> tableReport() |> head()
```

The eigenvalues TASSEL computes alongside the axes are the second report:

```{r, eval=TRUE, echo=TRUE}
myMDS |> tableReport("MDS_Eigenvalues_Datum")
```


## LD results
`linkageDiseq()` returns an `LDResults` object holding a single table of
pairwise statistics, which `tableReport()` returns:

```{r, eval=TRUE, echo=TRUE}
myLD <- linkageDiseq(
    tasObj  = myGT[, region("2", 228e6, 300e6)],
    ldType  = "All",
    verbose = FALSE
)

myLD |> tableReport() |> head()
```

Since there is only the one table, naming it and taking the default come
to the same thing:

```{r, eval=TRUE, echo=TRUE}
myLD |> reportNames()

identical(tableReport(myLD, "LD"), tableReport(myLD))
```

`"ALL"` behaves as it does everywhere else and returns a named list, of
one element.


## Functions that return R objects directly
Not every analysis is wrapped in a class. Several hand back an ordinary R
object, with nothing further to extract.

`seqDiversity()` returns a named `list` of `tibble` objects:

```{r, eval=TRUE, echo=TRUE}
myDiversity <- seqDiversity(myGT)

names(myDiversity)

myDiversity$Diversity[, 1:8] |> head()
```

`createTree()` returns an `ape` `phylo` object, ready for `plot()` or any
of the tree libraries that build on it:

```{r, eval=TRUE, echo=TRUE}
myTree <- createTree(myGT, clustMethod = "Neighbor_Joining")

class(myTree)
```

And `genomicPrediction()` returns a `tibble` of either accuracies or
predictions, depending on whether cross-validation was asked for:

```{r, eval=FALSE, echo=TRUE}
genomicPrediction(
    tasPhenoObj = myDataset,
    kinship     = myKin,
    doCV        = TRUE,
    kFolds      = 5,
    nIter       = 1
) |> head()
```
```
## # A tibble: 6 x 4
##   Trait Iteration  Fold Accuracy
##   <chr>     <dbl> <dbl>    <dbl>
## 1 EarHT         0     0    0.501
## 2 EarHT         0     1    0.376
## 3 EarHT         0     2    0.506
## 4 EarHT         0     3    0.595
## 5 EarHT         0     4    0.528
## 6 dpoll         0     0    0.785
```



# Reaching into Java
When something is not covered by an accessor, `javaRefObj()` returns the
pointer itself and `rJava`'s `$` operator calls TASSEL methods on it
directly:

```{r, eval=TRUE, echo=TRUE}
jGT <- myGT |> javaRefObj()

jGT

jGT$numberOfSites()

jGT$genotypeAsString(0L, 0L)
```

It works on the phenotype and dataset classes too:

```{r, eval=TRUE, echo=TRUE}
myPheno |> javaRefObj()
```

Two things to keep in mind. Java indices start at 0, so the call above
reads the first taxon at the first site. And this is an escape hatch rather
than a supported interface: what you reach is TASSEL's own API, which can
change between releases without `rTASSEL` changing with it. Prefer an
accessor where one exists.



# Superseded functions
Earlier versions of `rTASSEL` extracted data with dedicated functions
rather than base R generics. These still work but warn, and are scheduled
for removal in the next major release:

| Deprecated                       | Use instead                          |
|----------------------------------|--------------------------------------|
| `getPhenotypeDF()`               | `as.data.frame()`                    |
| `getSumExpFromGenotypeTable()`   | `as(x, "SummarizedExperiment")`      |
| `as.matrix()` on a `TasselGenotypePhenotype` | `as.matrix()` on a `TasselGenotype` or `TasselGenomicDataset` |

`getSumExpFromGenotypeTable()` built a `SummarizedExperiment` from the
dosage matrix and the position list, which is what the `as()` coercion
shown above now does.
