---
title: "Installing rTASSEL"
author: "Brandon Monier"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
    BiocStyle::html_document:
      toc: true
      fig_caption: true
      toc_float: true
      number_sections: false
vignette: >
    %\VignetteIndexEntry{Installing rTASSEL}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
    %\usepackage{float}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    fig.path='figure/graphics-',
    cache.path='cache/graphics-',
    fig.align='center',
    external=TRUE,
    echo=TRUE,
    warning=FALSE
    # fig.pos="H"
)
```

# Prerequisite - installing rJava

Since TASSEL is written primarily in Java, a Java JDK will need to be
installed on your machine. Additionally, for R to communicate with Java,
the R package [`rJava`](https://www.rforge.net/rJava/) will need to be
installed. In order to use `rTASSEL`, ensure that you have:

-   A `JDK` (Java Development Kit $\geq$ `8`) installed on your system.
-   Your system environment variable `JAVA_HOME` is configured
    appropriately and points to your `JDK` of choice. This will usually
    be included in your PATH environment variable as well. Options and
    system environmental variables that are available from R can be seen
    with `Sys.getenv()` and more specifically
    `Sys.getenv("JAVA_HOME")`.

**NOTE**: If you are using a UNIX system (e.g. Ubuntu) and are experiencing 
issues, you may need to reconfigure R with Java. To perform this, open a 
terminal and enter the command:

    R CMD javareconf

You may need to have root privileges when performing this so you may
need to add `sudo` to the prior command.

If you need additional steps on how to perform these actions, detailed
information can be found using the following links, depending on your
OS:

-   [Linux](https://datawookie.netlify.com/blog/2018/02/installing-rjava-on-ubuntu/)
-   [macOS](https://zhiyzuo.github.io/installation-rJava/)
-   [Windows](https://cimentadaj.github.io/blog/2018-05-25-installing-rjava-on-windows-10/installing-rjava-on-windows-10/)



# Install from GitHub

## Building with vignettes

After you have `rJava` up and running on your machine, install `rTASSEL`
by installing the source code from our GitHub repository using the
`devtools` package. Here, we show how you can install the package and build
vignettes locally:

```{r, eval=FALSE, echo=TRUE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github(
    repo = "maize-genetics/rTASSEL",
    ref = "master",
    build_vignettes = TRUE,
    dependencies = TRUE
)
```

The `dependencies = TRUE` parameter will have to be set if you do not have the
suggested packages described in the
[`DESCRIPTION`](https://github.com/maize-genetics/rTASSEL/blob/master/DESCRIPTION) 
file of this package.


## Building without vignettes

If you wish to **not** build vignettes, the prior method can be simplified
as shown below:

```{r, eval=FALSE, echo=TRUE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("maize-genetics/rTASSEL")
```



# Loading `rTASSEL`

After source code has been compiled, the package can be loaded using:

```{r, eval=TRUE, echo=TRUE}
library(rTASSEL)
```

Or, if you want to use a function without violating your environment you
can use `rTASSEL::<function>`, where `<function>` is an `rTASSEL`
function.



# Running from Docker

If you wish to run a containerized version of `rTASSEL`, we also have a 
[Docker image](https://hub.docker.com/repository/docker/maizegenetics/rtassel)
available. This can be retrieved from DockerHub using the following command:

```
docker pull maizegenetics/rtassel:latest
```

## With the terminal

Once downloaded, you can run `rTASSEL` from a terminal window:

```
docker run --rm -ti maizegenetics/rtassel R
```


## With RStudio Server

This image also contains an RStudio Server instance. To run this, you will
need to publish the container's port(s) to the host (`-p`). For example:

```
docker run --rm -ti -p 8787:8787 maizegenetics/rtassel
```

From here, you can go to `localhost:8787` on a web browser and enter a:

* Username (by default, this will be `rstudio`)
* Password (this will be a randomly generated password displayed in the 
  terminal output)


# Setting `rTASSEL`/Java memory

## Local overview
Since `rTASSEL` leverages the TASSEL 5 Java API via the `rJava` package, 
it is important to allocate sufficient memory to the Java Virtual Machine (JVM) 
before it starts. This is done using the `options(java.parameters = "-Xmx...")` 
command in R, which sets JVM parameters such as the maximum heap size (e.g., 
`-Xmx4g` for **4 GB**). The reason this must be set _before_ loading `rTASSEL` is
because the JVM can only be configured at startup. Once initialized, its memory 
settings cannot be changed without restarting the R session. This becomes 
especially important when working with large datasets or computationally 
intensive method calls, which can quickly exceed the default memory allocation 
and lead to `OutOfMemoryError`s. By increasing the available heap space 
proactively, we ensure that Java operations can be performed efficiently and 
without interruption due to memory constraints.

In short, if you are loading large genotype datasets and/or phenotype data, it
is adamant that you specify the memory allocated _**before**_ loading the
`rTASSEL` package via the `options()` function:

```{r, echo=TRUE, eval=FALSE}
# Allocate 4 GB of memory to the JVM
options(java.parameters = "-Xmx4g")

# Load rTASSEL
library(rTASSEL)
```


## Running `rTASSEL` on RStudio Server
Certain instances of RStudio Server on computing clusters can override what you 
specify in the prior example (i.e., running the `options()` function before 
loading `rTASSEL`) due to when the JVM is initialized and the `options()`
function is called in the recently initialized R session. If the JVM is 
initialized, any value provided to the `java.parameters` key in the `options()` 
call **will be silently ignored**. To prevent this from happening, make sure to
set up a `.Rprofile` configuration file with the aforementioned `options()` call:

```
## Example .Rprofile entry

# Allocate 4 GB of memory to the JVM
options(java.parameters = "-Xmx4g")
```

Since setting up a `.Rprofile` configuration are out of scope for this package,
please refer to [Posit's excellent write up on the subject](https://docs.posit.co/ide/user/ide/guide/environments/r/managing-r.html#rprofile).


## Helpful tips

### Verify memory has been set
If you are running into `OutOfMemory` exceptions, verify if you have specified
enough memory via the prior `options()` call. This can help verify if you have 
properly set enough memory at startup. By default, `rJava` will allocate 
**500 MB (0.5 GB)** of memory to your session. At any time during your R session
you can report the total memory allocated using a couple of `rJava` calls:

```{r, echo=TRUE, eval=FALSE}
# Call Java Runtime class
runtime  <- rJava::.jcall("java/lang/Runtime", "Ljava/lang/Runtime;", "getRuntime")

# Get total memory allocation (reported in gigabytes [GB])
gbConv <- 1024^3 # e.g. ~1e9 (billion) bytes
totMem <- round(round(rJava::.jcall(runtime, "J", "maxMemory") / gbConv, 3))

# Show in console
totMem
```

If the `java.parameters` value in `options()` was set up properly, the value 
specified in `totMem` should be the same value you specify in `options()`.

### Ensure enough memory is allocated
In most instances, genotype data will be the main determinant of how much
memory you should allocate to the JVM. In most cases, the amount of memory
you should allocate for genotype data is at least:

```
(# taxa) * (# sites) * (1 byte)
```

For example, if you have genotype data consisting of **250** taxa and **3000** 
sites, this would be **750000** bytes or **0.75** megabytes (MB).


# Prior issues and possible resolutions

## Problems installing rJava on macOS with M1 CPU architecture

If you are running into issues with installing `rJava` using the newer
Mac chip architecture, Oracle JDK currently (as of writing this) does not 
work. Consider an alternative JDK source such as [OpenJDK](https://openjdk.org/)
or [Azul JDK](https://www.azul.com/downloads/?version=java-8-lts&package=jdk).

More detailed information about a possible workaround can be found in this
[Stack Overflow post](https://stackoverflow.com/questions/67849830/how-to-install-rjava-package-in-mac-with-m1-architecture).


## Problems installing if you have both 32- and 64-bit architecture installed for R

If you are using a machine that has **both architectures
installed for R**, you might run into problems pulling code using
`devtools`. If this is the case, one solution would be to add the
parameter `--no-multiarch` option in `INSTALL_opts`. This will force
building the package for your currently running R version:

```{r, eval=FALSE, echo = TRUE}
devtools::install_github(
    repo = "maize-genetics/rTASSEL",
    ref = "master",
    build_vignettes = FALSE,
    INSTALL_opts = "--no-multiarch"
)
```


## Problems with `rJava` if you have upgraded Java

On macOS: if you previously had `rJava` working through RStudio, then
you upgraded your Java and it now longer works, try the following:

At the command line type:

    R CMD javareconf

Then check for a left over symbolic link via:

    ls -ltr /usr/local/lib/libjvm.dylib

If the link exists, remove it, then create it fresh via these commands:

    rm /usr/local/lib/libjvm.dylib
    sudo ln -s $(/usr/libexec/java_home)/lib/server/libjvm.dylib /usr/local/lib

You should now be able to enter RStudio and setup `rJava`.




