---
title: "Installing rTASSEL"
author: "Brandon Monier"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
    BiocStyle::html_document:
      toc: true
      fig_caption: true
      toc_float: true
      number_sections: false
vignette: >
    %\VignetteIndexEntry{Installing rTASSEL}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
    %\usepackage{float}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    fig.path='figure/graphics-',
    cache.path='cache/graphics-',
    fig.align='center',
    external=TRUE,
    echo=TRUE,
    warning=FALSE
    # fig.pos="H"
)
```

# Prerequisite - installing rJava

Since TASSEL is written primarily in Java, a Java JDK will need to be
installed on your machine. Additionally, for R to communicate with Java,
the R package [`rJava`](https://www.rforge.net/rJava/) will need to be
installed. In order to use `rTASSEL`, ensure that you have:

-   A `JDK` (Java Development Kit $\geq$ `8`) installed on your system.
-   Your system environment variable `JAVA_HOME` is configured
    appropriately and points to your `JDK` of choice. This will usually
    be included in your PATH environment variable as well. Options and
    system environmental variables that are available from R can be seen
    with `Sys.getenv()` and more specifically
    `Sys.getenv("JAVA_HOME")`.

**NOTE**: If you are using a UNIX system (e.g. Ubuntu) and are experiencing 
issues, you may need to reconfigure R with Java. To perform this, open a 
terminal and enter the command:

    R CMD javareconf

You may need to have root privileges when performing this so you may
need to add `sudo` to the prior command.

If you need additional steps on how to perform these actions, detailed
information can be found using the following links, depending on your
OS:

-   [Linux](https://datawookie.netlify.com/blog/2018/02/installing-rjava-on-ubuntu/)
-   [macOS](https://zhiyzuo.github.io/installation-rJava/)
-   [Windows](https://cimentadaj.github.io/blog/2018-05-25-installing-rjava-on-windows-10/installing-rjava-on-windows-10/)



# Install from GitHub

After you have `rJava` up and running on your machine, install `rTASSEL`
by installing the source code from our GitHub repository using the
`devtools` package:

```{r, eval=FALSE, echo=TRUE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github(
    repo = "maize-genetics/rTASSEL",
    ref = "master",
    build_vignettes = TRUE,
    dependencies = TRUE
)
```

After source code has been compiled, the package can be loaded using:

```{r, eval=TRUE, echo=TRUE}
library(rTASSEL)
```

Or, if you want to use a function without violating your environment you
can use `rTASSEL::<function>`, where `<function>` is an `rTASSEL`
function.


# Running from Docker

If you wish to run a containerized version of `rTASSEL`, we also have a 
[Docker image](https://hub.docker.com/repository/docker/maizegenetics/rtassel)
available. This can be retrieved from DockerHub using the following command:

```
docker pull maizegenetics/rtassel:latest
```

Once downloaded, you can run `rTASSEL` from an R terminal:

```
docker run --rm -ti maizegenetics/rtassel R
```

Or you can start an RStudio Server instance. For example:

```
docker run --rm -ti -p 8787:8787 maizegenetics/rtassel
```



# Prior issues and possible resolutions

## Problems with `rJava` if you have upgraded Java

On macOS: if you previously had `rJava` working through RStudio, then
you upgraded your Java and it now longer works, try the following:

At the command line type:

    R CMD javareconf

Then check for a left over symbolic link via:

    ls -ltr /usr/local/lib/libjvm.dylib

If the link exists, remove it, then create it fresh via these commands:

    rm /usr/local/lib/libjvm.dylib
    sudo ln -s $(/usr/libexec/java_home)/lib/server/libjvm.dylib /usr/local/lib

You should now be able to enter RStudio and setup `rJava`.


## Problems installing if you have both 32- and 64-bit architecture installed for R

**NOTE:** If you are using a machine that has both architectures
installed for R, you might run into problems pulling code using
`devtools`. If this is the case, one solution would be to add the
parameter `--no-multiarch` option in `INSTALL_opts`. This will force
building the package for your currently running R version:

```{r, eval=FALSE, echo = TRUE}
devtools::install_github(
    repo = "maize-genetics/rTASSEL",
    ref = "master",
    build_vignettes = FALSE,
    INSTALL_opts = "--no-multiarch"
)
```


